---
title: "Lab 3 Tutorial"
author: "Janette Avelar & Anwesha Guha"
date: "1/25/2022"
output:
  ioslides_presentation:
    widescreen: yes
    smaller: no
  html_document:
    df_print: paged
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro
This week we will go over calculating probability from a standard normal distribution.

Note: **All material is taken from Dr. Zopluoglu's 2021 EDUC 614 materials and adapted for this year's course.**

We have adapted it to slide format using R.  

*Want to learn how we created R slide presentations?* [This tutorial](https://www.r-bloggers.com/2019/09/mastering-r-presentations/) *is helpful.*

# Calculating Probability from a Standard Normal Distribution

## Density for z-score

To compute density for a z-score, you can compute `dnorm()` from the base package. Below is the code to reproduce density calculations from Lecture 3a.

```{r}
dnorm(x = -1)

dnorm(x = 0)
  
dnorm(x = 1)
  
dnorm(x = 2)
```

## Lower-tail Probability

Lower-tail probability refers the probability of a random variable X taking the value x or less. In other words, it refers to the area under the normal curve on the left side of normal distribution.
  
Cumulative probability is also another term being used for lower-tail probability. 
  
You can compute this by using the `pnorm()` function from the base package.

## Examples

Suppose, you want to compute the probability of z scores being 0 or less than 0.
  
**P(z < 0) = ?**

```{r}
pnorm(0, lower.tail = TRUE)
```

##
What about probability of z score being -1 or less than -1?

P(z < -1) = ?

```{r}
pnorm(-1, lower.tail = TRUE)
```

##
Or probability of z score being 2 or less than 2?

P(z < 2) = ?

```{r}
pnorm(2, lower.tail = TRUE)
```

## Upper-tail Probability

Upper-tail probability refers the probability of a random variable X taking the value x or more. In other words, it refers to the area under the normal curve on the right side of normal distribution.

You can compute this by using `pnorm()` from the base package, and setting `lower.tail = FALSE`.

## Examples

Suppose you want to calculate the probability of z score being -1 or more than -1.

**P(z > -1) = ?**

```{r}
pnorm(-1, lower.tail = FALSE)
```

##
What about the probability of z score being 2 or more than 2?

**P(z > 2) = ?**
  
```{r}
pnorm(2, lower.tail = FALSE)
```

## Probability between scores
If you want to compute the probability between two z-scores, then you need to get creative.  
  
Suppose you want to compute the probability of z score being between -1 and 2.  
  
You can do this by subtracting P(z<-1) from P(z<2).
  
```{r}
pnorm(2, lower.tail = TRUE) -  pnorm(-1, lower.tail = TRUE)
```

## `qnorm()`

Another useful function is `qnorm()`. Consider this function as an inverse of `pnorm()`.
  
- `pnorm()` returns the cumulative probability given a z-score in a normal distribution
  
- `qnorm()` returns the z-score given the cumulative probability.

## Example

Suppose you want to know that which z-score has a cumulative probability of 0.6.
  
```{r}
qnorm(0.6,lower.tail=TRUE)
```

# Q-Q Plots

## Import the data

Like in our previous tutorial where we learned how to create histograms and boxplots, in order to create a q-q plot our first step is to import the dataset we'd like to use.

```{r}
Add <- read.csv(file   = "/Users/janetteavelar/Desktop/Add.csv",
                header = TRUE)
```
Remember your working directory goes in the first line of code. You can set it manually, or use `getwd()` to find your file path.

*Note:* If you choose to import your dataset using the point-and-click interface, be sure to mark **Yes** for Header to ensure your data loads in properly.

## Create the plot

We'll now use the `qqline()` function from the base package to draw a QQ plot for any variable of choice.

Let's look at GPA first.

```{r}
qqnorm(Add$GPA)
qqline(Add$GPA)  # This adds a diagonal line
```

## Another example

Now let's try another variable.

```{r}
qqnorm(Add$ADDSC)
qqline(Add$ADDSC)  # Add the diagonal line
```

