---
title: "Lab 7 Tutorial"
author: "Janette Avelar & Anwesha Guha"
date: "1/25/2022"
output:
  ioslides_presentation:
    widescreen: yes
    smaller: no
  html_document:
    df_print: paged
  beamer_presentation: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chi-square Test of Independence

## Intro
This week we will go over conducting a chi-square test of independence. This tutorial will use the same example that was used in class, offering you step-by-step explanation for the analysis that you may reference in the future.

Note: **All material is taken from Dr. Zopluoglu's 2021 EDUC 614 materials and adapted for this year's course.**

We have adapted it to slide format using R.  

*Want to learn how we created R slide presentations?* [This tutorial](https://www.r-bloggers.com/2019/09/mastering-r-presentations/) *is helpful.*

## Importing Data

Our first step is to import the data we need. We'll use the `Personality` dataset for this walkthrough, which you can access via Canvas. We'll also use the `factor()` and `str()` functions to ensure the data is in the proper format to run an ANOVA.  

```{r}
pers <- read.csv(file   = "/Users/janetteavelar/Desktop/personality.csv",
                header = TRUE)
```

Our data consists of 2 columns, the first indicating personality type and the second indicating color preference. We'll want to convert these variables so they make sense for our analysis